The Private School Network: Recruiting Visits to Private High Schools by Public and Private Universities

### 2023-09-06 r_r/front_end.Rmd
draft changes front end

## Scholarship on Nonresident Enrollment

We situate this study vis-a-vis scholarship on nonresident enrollment at public research universities, which lies within the broader literature on "privatization" [CITE]. Against the backdrop of many states disinvesting in higher education, the privatization literature examines the ways that public universities start behaving like private universities by emphasizing revenue generation and the pursuit of prestige in research and enrollment management behaviors. The so-called "iron triangle" of enrollment management states that universities pursue the three broad enrollment goals of access, academic profile, and revenue generation [CITE]. "Because resources are scarce, the imagery of the iron triangle suggests that the pursuit of one enrollment goal may involve tradeoffs with other goals. Scholarship on privatization expressed concern that the public research universities were emphasizing tuition revenue and rankings and deemphasizing access for state residents.

@@

How did public research universities go about attracting all these nonresident students. One mechanism is financial aid. @RN2535 shows that nonresident students are less sensitive to sticker price than resident students. Because nonresident prospects may be considering several options -- in-state, out-of-state, public, private -- they are sensitive to the generosity of institutional aid offers [CITE]. For example, @RN3762 analyze the effect of a University of Iowa institutional merit aid program for nonresident admits that offered higher awards for admits with higher academic achievement. Regression discontinuity analyses found that the program increased the probability of enrollment, particularly for high achieving students. These results suggest that high-achieving nonresident students face many good opportunities and require substantial institutional aid to be convinced to enroll.

Earlier in the "enrollment funnel," universities utilize marketing and recruiting interventions to attract nonresident applicants. One widely utilized intervention is off-campus recruiting visits by admissions counselors to local high schools. @RN3519 found that colleges value this visits as a means of meeting prospective students and maintaining relationships with guidance counselors at feeder schools. @RN4758 analyzed off-campus recruiting visits by 15 public research universities, finding that 12 of these universities made more visits to out-of-state
high schools than in-state high schools. These out-of-state visits focused on public schools in affluent, predominantly white communities. Although @RN4758 focused on visits to public high schools, they noted a surprising number of visits to out-of-state private high schools. However, we know little about this private school visits, for example, which sorts of schools are they visiting and at what scale.

## Scholarship on Linkages Between Private Schools and Colleges

A separate thread of scholarship analyzes linkages between private schools and colleges. Nationwide, dddd

TEXT YOU WROTE ABOUT MURNANE AND REARDON
In particular, Murnane & Reardon (2018) used data from several national surveys to analyze long-term trends in private school enrollment by family income. The percentage of American (elementary school) students attending private school declined from 15% in the With respect to family income, the private school enrollment rate of high-income families (around 17%) and low-income families (around 4%) remained stable over time, but the enrollment rate for middle-income families declined from 12% in 1968 to 6% in 2013. These declines were substantially driven by declines in the number of students attending Catholic
schools, which represented 89% of private school enrollment in 1965 and 42% of enrollment in 2013. Private school enrollment rates are higher for families located in urban areas than those in suburban areas. With respect to race, in 2013, 11% of white students attended private school compared to 5% of Black students, and 3% of Hispanic students. However, for high income families, rates of private school enrollment do not differ substantially across race. In the South, there has been a long-term increase in private school enrollment, driven
by growing enrollment in Christian private schools by middle-income families and growing enrollment in nonsectarian private schools by high-income families.
mid-1960s to 10% in the mid-1970s, and declined gradually from 11% in 1999 to 9% in 2015.

- I think good to say that private schools make up about 9% of k-12 (or is it primary) enrollment
The majority of research

>>>>

## Nonresident Enrollment at Public Universities

We stiuate this study vis-a-vis scholarship on enrollment management behavior by public research universities and particularly the pursuit of affluent nonresident students, a population also targeted by selective private universities. The so-called "iron triangle" of enrollment management states that universities pursue the broad enrollment goals of access, academic profile, and revenue generation [@RN3524]. Historically, public research universities prioritize enrollment of high-performing state residents, often with an emphasis on social mobility for students who cannot afford private universities [@RN3608]. By contrast, selective private universities typically prioritize academic profile and revenue generation, with funds generated from tuition and donation feeding the arms race for prestige [@RN3522; @RN1549]. Because resources are scarce, the imagery of the iron triangle suggests that the pursuit of one enrollment goal may involve trade-offs with other goals. @RN2772[p. 221] state that “most enrollment management policies\ldots do not advance all three objectives; instead they lead to gains in some areas and declines in others.” In the 2000s, scholarship on privatization argued the public research universities were emphasizing tuition revenue and rankings and de-emphasizing access for state residents [@RN1577; @RN1118].
<!-- PREVIOUS PARAGRAPH
We situate this study vis-a-vis scholarship on nonresident enrollment at public research universities, which lies within the broader literature on "privatization" [CITE]. Against the backdrop of many states disinvesting in higher education, the privatization literature examines the ways that public universities start behaving like private universities by emphasizing revenue generation and the pursuit of prestige in research and enrollment management behaviors. The so-called "iron triangle" of enrollment management states that universities pursue the three broad enrollment goals of access, academic profile, and revenue generation [CITE]. "Because resources are scarce, the imagery of the iron triangle suggests that the pursuit of one enrollment goal may involve tradeoffs with other goals. Scholarship on privatization expressed concern that the public research universities were emphasizing tuition revenue and rankings and deemphasizing access for state residents.-->

@@

How did public research universities go about attracting all these nonresident students. One mechanism is financial aid. Many public research universities have developed institutional aid policies for nonresident admits, typically offering larger awards for higher achieving students [@RN3762; @RN2287; @RN1791].<!--@RN2535 shows that nonresident students are less sensitive to sticker price than resident students. Because nonresident prospects may be considering several options -- in-state, out-of-state, public, private -- they are sensitive to the generosity of institutional aid offers [CITE]. For example, @RN3762 analyze the effect of a University of Iowa institutional merit aid program for nonresident admits that offered higher awards for admits with higher academic achievement. Regression discontinuity analyses found that the program increased the probability of enrollment, particularly for high achieving students. These results suggest that high-achieving nonresident students face many good opportunities and require substantial institutional aid to be convinced to enroll. -->

Earlier in the "enrollment funnel," universities utilize marketing and recruiting interventions to attract nonresident applicants. One widely utilized intervention is off-campus recruiting visits by admissions counselors to local high schools. @RN4758 analyzed off-campus recruiting visits to public high schools by 15 public research universities. Recruiting visits to in-state high schools tended exhibit modest racial and socioeconomic bias. However, 12 of the 15 universities made more visits to out-of-state high schools than in-state high schools. These out-of-state visits focused on public schools in affluent, predominantly white communities. @RN4934 conducted geo-spatial analyses of out-of-state recruiting visits to Los Angeles and Dallas by four public research universities. She found that universities engage in "recruitment redlining -- the circuitous avoidance of predominantly Black and Latinx communities along recruiting visit paths" (p. X). Although @RN4758 and @RN4934 focused on visits to public high schools, they noted a surprising number of visits to out-of-state private high schools. 
<!-- However, we know little about this private school visits, for example, which sorts of schools are they visiting and at what scale. -->

## Linkages Between Private Schools and Private Universities

Disconnected from scholarship on nonresident enrollment at public universities, scholarship from economics and sociology analyzes linkages between private schools and colleges.

@RN4504<!-- use data from several national surveys to --> set the stage by analyzing long-term trends in private school enrollment. The percentage of American (elementary school) students attending private school declined from 15% in the mid-1960s to 10% in the mid-1970s, and declined gradually from 11% in 1999 to 9% in 2015. With respect to family income, the private school enrollment rate of high-income families (around 17%) and low-income families (around 4%) remained stable over time, but the enrollment rate for middle-income families declined from 12% in 1968 to 6% in 2013. These declines were substantially driven by declines in the number of students attending Catholic schools, which represented 89% of private school enrollment in 1965 and 42% of enrollment in 2013. However, as Catholic school enrollment has declined, Catholic school students tend to be more affluent than they were in prior decades. In the South, there has been a long-term increase in private school enrollment, driven by growing enrollment in Christian private schools by middle-income families and growing enrollment in nonsectarian private schools by high-income families. With respect to race, in 2013, 11% of white students attended private school compared to 5% of Black students, and 3% of Hispanic students.
<!-- Private school enrollment rates are higher for families located in urban areas than those in suburban areas.
 However, for high income families, rates of private school enrollment do not differ substantially across race. -->

A robust empirical literature<!-- from sociology and economics  --> finds that attending private school positively affects college access and college selectivity [e.g.,] [@RN4642; @RN4596; @RN4634; @RN4599]. Scholarship in the 1990s often focused on the effects of attending Catholic schools [@RN4937; @RN786; @RN4936; @RN4938; @RN4939]. This is because Catholic schools enrolled the majority of private school students and because Catholic school students tended to be less affluent than those at other private schools. <!-- More recent scholarship tends to find that attending a nonsectarian private school has a larger effect on college access outcomes compared to Catholic schools [@RN4614;@RN4625].  --> DESCRIPTIVE STATISTICS FROM HSLS ABOUT PRIVATE SCHOOLS AND SELECTIVE PRIVATE SCHOOLS.

Scholarship has been particularly interested in the relationship between attending an elite private school and attending an elite private college [@RN4647; @RN4643; @RN4647; @RN4589]. @RN4722 acquired unprecedented data -- application files and admissions decisions linked to income tax records  -- to investigate why high income families are over-represented at "Ivy Plus" colleges. Conditioning on SAT and ACT scores, applications from high-income families (top 1%) received the same academic rating as applications from lower-income families. However, applications from high-income families tended to receive higher non-academic ratings (extracurricular activities, leadership traits) then applications from lower-income families. In turn, the higher non-academic ratings for high-income applicants was driven by the fact that these students tended to attend elite private high schools that devote substantial resources to teacher recommendations and guidance counselor letters.

Scholarship from sociology has examined organizational connections that catalyze the flow of students between private schools and selective private colleges [@RN4640; @RN4407; @RN4321; @RN3519].

@RN3519 provides an ethnography of the admissions office at a highly selective (but not elite) private liberal arts college. The College is sensitive about its position in U.S. News Rankings and enrollment priorities tend to focus on academic profile and revenue generation. @RN3519 highlights the relational function of off-campus recruiting visits, stating that "the College's reputation and the quality of its applicant pool are dependent upon its connections with high schools nationwide" (p. 54) Therefore, during the autumn "travel season," admissions officers visit selected high schools across the country "to spread word of the institution and maintain relationships with guidance counselors" (p. 53-54). The logic is that a guidance counselor who views a college favorably will steer students to the college. The College tended to visit the same "feeder" schools year after year because recruiting depends on long-term relationships with high schools. The high schools they visited tend to be affluent schools -- in particular, private schools -- that enroll high-achieving students who can afford tuition and have the resources and motivation to host a successful visit. 
<!-- Whereas @RN4402 highlights the effect of recruiting visits on inquiries and enrollees, findings from @RN3519 suggest that the College may have valued recruiting visits primarily as a means of maintaining relationships with guidance counselors. From this perspective, recruiting visits may affect outcomes such as inquiries, applications, and matriculation through their effect on high school guidance counselors. -->

@RN4321 analyzed recruiting from the perspective of an elite private boarding school in order to understand "how such schools continue to get comparatively under-qualified students into top colleges and universities" (p. 98). Guidance counselors at elite private schools capitalize on the fact that college admissions offices value credible information about which applicants will accept or decline an admissions offer because colleges are sensitive to their acceptance and yield rates. This desire for credible intelligence creates an opportunity for high school counselors to advocate on behalf of their students. This opportunity depends on guidance counselors having personal relationships with university admissions offices and on having small enough caseloads to advocate for each student individually. Guidance counselors tell admissions counselors which highly-sought-after applicants are likely to decline an offer, while lobbying for an applicant who relatively weak academic credentials, but has "character," and extracurricular activities the college needs, and who is sure to matriculate.

## Synthesis

Our review of two literatures -- scholarship on nonresident enrollment by public research universities and scholarship on linkages between private schools and private colleges -- suggests an opportunity to bridge and contribute to both literatures.

Historically, public universities received most of their revenue from state funding and enrollment goals prioritized opportunities for in-state students. As state funding became uncertain, public research universities began to prioritize tuition revenue and sought affluent, out-of-state students who paid high nonresident tuition price. Salazar et al shows that recruiting visits by public research universities out-of-state public schools in affluent, predominantly white communities. Resource dependence theory suggests that as public research universities become more tuition reliant, they will start behaving more like selective private universities in targeting private schools which tend to enroll affluent students that can to pay high tuition prices. However, prior research has not investigated visits to private schools. For example, which sorts of private schools are public research universities visiting and at what scale?

The majority of scholarship on linkages between private high schools and private universities focuses on elite high schools and elite universities [CITE]. We know much less about linkages between private schools and private universities that are selective but not elite. These universities are sensitive about their reputation. Despite being relatively affluent, they are much more resource constrained than Ivy Plus. Thus, they cannot pursue all enrollment goals with equal vigor. Prior research shows that selective but not elite colleges prioritize academic profile (as defined by US News Rankings) and tuition revenue, potentially at the expense of access. Many private high schools enroll students that satisfy both enrollment goals. However, prior scholarship provides little insight about which private schools have relationships with selective private universities.

We argue that a recruiting visit from a university to a private high school is an indicator of a relationship. Empirically, @RN3519 finds that off-campus recruiting visits are important for the maintenance of strong relationships between a college and a high school. Logically, the fact that the college made the effort to visit suggests that the college wants to enroll students from the high school. Similarly, the fact that the high school hosted the visit suggests that the high school likely views the college as a desirable destination for some of its students. Moreover, the presence of the recruiting visit suggests the probability of additional interactions (e.g., phone calls) between the organizations. Conceptualizing a recruiting visit as an indicator of a relationship between a college and a high school motivates the use of social network methods, which analyze the network defined by "network ties" (i.e., visits) between actors (universities and schools).

Because nonresident enrollment is a relatively new phenomenon for most public research universities [CITE], visits to out-of-state private schools are encroachments on territory that was previously the domain of selective private universities and in-state public flagships. These visits suggest that public research universities are competing with selective private universities for private school students. This may be direct competition for the same student or vying to be the second choice for students who are rejected by their top choice. By analyzing the network of recruiting visits, we can determine the extent to which public research universities are visiting the same or similar sets of private high schools. These analyses provide a novel lens into the nature of competition between public research universities and selective private universities. They also yield insight about which student bodies these universities want to enroll.
